---
title: "Boston Housing Regression Analysis"
author: "Abdullah Shahzad"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

This report presents a detailed regression analysis of the **Boston Housing** dataset.  
The goal is to investigate how various housing characteristics affect the **median value of owner-occupied homes** (`medv`) in different suburbs of Boston.

We will:

1. Explore and visualize the dataset.
2. Identify key predictors of `medv`.
3. Fit and evaluate regression models (linear and polynomial).
4. Interpret the results and insights.

The dataset comes from the **MASS** package and contains 506 observations with 14 variables.

---

# Load Libraries and Data

```{r setup, message=FALSE, warning=FALSE}
library(MASS)      # For Boston dataset
library(tidyverse) # Data manipulation and visualization
library(car)       # For VIF calculations
library(corrplot)  # For correlation plots
library(caret)     # For train-test splitting
```

```{r load-data}
# Load the dataset
data("Boston")

# Quick look at the data
glimpse(Boston)
summary(Boston)

# Check for missing values
colSums(is.na(Boston))
```

---

# Exploratory Data Analysis

## Correlation with Target Variable

```{r correlation}
corr_matrix <- cor(Boston)
corrplot(corr_matrix, method = "color", tl.cex = 0.7)

sort(corr_matrix[, "medv"], decreasing = TRUE)
```

**Observation:**  
- `rm` (average number of rooms) has a strong positive correlation with `medv`.  
- `lstat` (% lower status) has a strong negative correlation.  

---

## Distribution of Median Home Value

```{r medv-distribution}
Boston %>% 
  ggplot(aes(x = medv)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  geom_vline(aes(xintercept = mean(medv)),
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = mean(Boston$medv) + 5, y = 60,
           label = paste("Mean:", round(mean(Boston$medv), 2))) +
  labs(title = "Distribution of Median Home Value", x = "medv", y = "Count")
```

```{r medv-boxplot}
Boston %>% 
  ggplot(aes(y = medv)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of Median Home Value")
```

---

## Scatterplots for Key Variables

```{r scatterplots}
pairs(~ rm + lstat + medv,
      data = Boston,
      panel = panel.smooth)
```

**Insights:**
- `rm` vs `medv`: Strong linear relationship.
- `lstat` vs `medv`: Downward curve, suggesting non-linearity.
- `lstat` vs `rm`: Negative correlation.

---

# Model Building

## Model 1: Simple Multiple Linear Regression

```{r model1}
model1 <- lm(medv ~ rm + lstat, data = Boston)
summary(model1)

# Plot predicted vs actual
pred <- predict(model1)
actual <- Boston$medv
plot(pred, actual, pch = 20,
     xlab = "Predicted Values", ylab = "Actual Values")
abline(0, 1, col = "red", lwd = 2)
```

**Observation:** Model explains ~64% of variance (R²), but struggles with extreme values.

---

## Model 2: Adding More Predictors

```{r model2}
model2 <- lm(medv ~ rm + lstat + ptratio + tax + black + crim, data = Boston)
summary(model2)

vif(model2)
```

**Observation:**  
- Some predictors (e.g., `tax`, `crim`) are not statistically significant.
- Multicollinearity is present but manageable.

---

## Model 3: Removing Insignificant Predictors

```{r model3}
model3 <- lm(medv ~ rm + lstat + ptratio + black, data = Boston)
summary(model3)
```

---

# Polynomial Regression for Non-Linearity

Since `lstat` shows curvature in its relationship with `medv`, we add a squared term.

```{r polymodel}
polymodel <- lm(medv ~ rm + lstat + I(lstat^2) + ptratio + black, data = Boston)
summary(polymodel)

vif(polymodel)
```

```{r lstat-plot}
lstat_vals <- seq(min(Boston$lstat), max(Boston$lstat), length.out = 100)
rm_mean <- mean(Boston$rm)
ptratio_mean <- mean(Boston$ptratio)
black_mean <- mean(Boston$black)

newdata <- data.frame(
  rm = rm_mean,
  lstat = lstat_vals,
  ptratio = ptratio_mean,
  black = black_mean
)

predicted_medv <- predict(polymodel, newdata = newdata)

plot(Boston$lstat, Boston$medv, pch = 16, col = "gray",
     xlab = "LSTAT (% lower status)", ylab = "MEDV",
     main = "Effect of LSTAT on MEDV (Squared Term)")
lines(lstat_vals, predicted_medv, col = "blue", lwd = 2)
```

---

## Polynomial on `rm` as Well

```{r poly2}
poly2 <- lm(medv ~ rm + I(rm^2) + lstat + I(lstat^2) + ptratio + black, data = Boston)
summary(poly2)
```

```{r rm-plot}
rm_vals <- seq(min(Boston$rm), max(Boston$rm), length.out = 100)
lstat_mean <- mean(Boston$lstat)

newdata <- data.frame(
  lstat = lstat_mean,
  rm = rm_vals,
  ptratio = ptratio_mean,
  black = black_mean
)

predicted_medv <- predict(poly2, newdata = newdata)

plot(Boston$rm, Boston$medv, pch = 16, col = "gray",
     xlab = "RM (avg rooms)", ylab = "MEDV",
     main = "Effect of RM on MEDV (Squared Term)")
lines(rm_vals, predicted_medv, col = "blue", lwd = 2)
```

---

# Model Diagnostics

```{r diagnostics}
par(mfrow = c(2, 2))
plot(poly2)
```

**Observations:**
- Residuals show less curvature compared to earlier models.
- Still some outliers at high `medv` values.

---

# Train-Test Evaluation

```{r train-test}
set.seed(123)
train_index <- createDataPartition(Boston$medv, p = 0.8, list = FALSE)
train_data <- Boston[train_index, ]
test_data <- Boston[-train_index, ]

model_train <- lm(medv ~ rm + I(rm^2) + lstat + I(lstat^2) + ptratio + black,
                  data = train_data)

pred_train <- predict(model_train, newdata = train_data)
rmse_train <- sqrt(mean((pred_train - train_data$medv)^2))

pred_test <- predict(model_train, newdata = test_data)
rmse_test <- sqrt(mean((pred_test - test_data$medv)^2))

rmse_train
rmse_test
```

---

# Conclusion

- `rm` and `lstat` are the strongest predictors of `medv`.
- Polynomial terms for `lstat` and `rm` improve model fit.
- Final polynomial multiple regression reduces error and captures curvature in the data.
- Caution: Model still struggles with extreme `medv` values, indicating potential need for robust regression or transformation.
